Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-02-20 brms_meeg.qmd
adding preliminary results with actual MEG data

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive processes. Typically, researchers are interested in testing whether and when such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics through time followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset or offset of effects (e.g., differences across conditions or groups). Here, we introduce a *model-based* approch for analysing M/EEG timeseries such as ERPs or timecourses of decoding performance and their differences across conditions or groups. This approach relies on Bayesian generalised additive multilevel models, which output the posterior probabilility of the effect being above 0 (or above chance) at every timestep, while naturally taking into account the temporal dependencies and between-subject variability present in such data. Using both simulation and actual EEG data, we show that the proposed approach largely outperforms conventional methods in determining both the onset and offset of M/EEG effects (e.g., ERPs difference, decoding performance), producting more precise and more reliable estimates. We provide an R package implementing the approach and illustrate how to integrate it into M/EEG statistical pipelines in MNE-Python."
keywords: [EEG, MEG, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

@@

@@

From @dimigen2021: Recently, spline regression has been applied to ERPs [e.g., @hendrix2017; @kryuchkova2012]... GAMMs for EEG data [@abugaber2023; @meulman2015]...

@@

Using Bayes factors for decoding performance [@teichmann2022]...

@@

See for instance these tutorials [@sóskuthy2017; @winter2016] or application to phonetic data [@wieling2018] or this introduction [@baayen2020] or these reference books [@hastie2017; @wood2017]... application to pupillometry [@vanrij2019]... GAMLSS for neuroimaging data [@dinga2021]...

@@

$$
\begin{aligned} 
y_{i} &\sim \mathrm{EF}\left(\mu_{i}, \phi\right)\\
g\left(\mu_i\right) &=  \underbrace{A_{i} + \mathbf{X}_{i} \gamma}_{\text{parametric part}} + \underbrace{\sum_{j=1}^{J} f_{j}\left(x_{ij}\right)}_{\text{non-parametric part}}
\end{aligned}
$$

where $y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)$ denotes that the observations $y_{i}$ are distributed as some member of the exponential family of distributions (e.g., Gaussian, Gamma, Beta, Poisson) with mean $\mu_{i}$ and scale parameter $\phi$; $g(\cdot)$ is the link function, $A$ is an offset, $\mathbf{X}_{i}$ is the $i$th row of a parametric model matrix, $\gamma$ is a vector of parameters for the parametric terms, $f_{j}$ is a smooth function of covariate $x_{j}$. The smooth functions $f_{j}$ are represented in the model via penalised splines basis expansions of the covariates, that are a weighted sum of basis functions:

@@

@@

Now describe the Bayesian GAMM... Proper inclusion of varying/random effects in the model specification protects against overly wiggly curves [@baayen2020]...

@@

Following the approach used by @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see details and code in @yeung2004]. As in @rousselet_using_2025, the signal was generated from truncated Gaussian with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated such data for a group of 20 participants with 50 trials per participant and condition (@fig-eeg).

@@

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms`, that is, weakly informative priors. We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects.

@@

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "cr", k = 10, by = condition),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/gam.rds"
    )
```

@@

# fitting the GAMM with factor smooth (per participant)
gamm <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition +
        s(time, bs = "cr", k = 10, by = condition) +
        s(time, participant, bs = "fs"),
    data = raw_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/gamm.rds"
    )
```

@@

```{r fig-post-prob-ratio, echo = FALSE, out.width = "75%", fig.asp = 0.75, fig.cap = "Ratio of posterior probability according to the GAM. Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.05
threshold <- 10

@@

# printing the results
# bind_rows(exceeding_times_gam, exceeding_times_gp) %>%
#     mutate(across(.cols = where(is.numeric), .fns = ~round(.x, 2) ) ) %>%
#     relocate(model, .before = cluster_onset)
```

@@

Next we fit a hierarchical/multilevel GAM using summary statistics of ERPs (mean and SD) at the participant level (similar to what is done in meta-analysis).

@@

# fitting the GAM
meta_gam <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 10, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/meta_gam.rds"
    )
```

@@

Assessing the reliability of the proposed approach using some sort of split-half reliability [e.g., @rosenblatt2018]? Using the MEG decoding results of Nalborczyk et al. (in preparation) in 33 participants:

@@

```{r reliability, echo = FALSE, eval = FALSE}
# library(caret)
# see https://github.com/topepo/caret/
# https://topepo.github.io/caret/data-splitting.html
# and createDataPartition()
# and createFolds()
# trainIndex <- createDataPartition(
#     y = summary_df$participant,
#     p = 0.5,
#     list = FALSE,
#     times = 10
#     )

# subsetting the data
# summary_df_train <- summary_df[trainIndex, ]
# summary_df_test  <- summary_df[-trainIndex, ]

# getting unique participants
unique_participants <- unique(summary_df$participant)

# getting the number of participants
num_participants <- length(unique_participants)

# ensuring participants are randomly shuffled
shuffled_participants <- sample(unique_participants)

# defining the number of splits
num_splits <- 10

# creating a dataframe where each participant appears in every split
split_df <- crossing(participant = unique_participants, split_number = 1:10)

# assigning alternating groups (A or B) within each split for each participant
# split_df <- split_df %>% mutate(group = rep(c("A", "B"), length.out = n() ) )
split_df <- split_df %>%
    group_by(split_number) %>%
    # shuffling groups for each split
    mutate(group = sample(rep(c("A", "B"), length.out = n() ) ) ) %>%
    ungroup()

# sanity checks
# table(split_df$group)
# split_df %>%
#     ggplot(aes(x = group) ) +
#     geom_histogram(stat = "count") +
#     facet_wrap(~participant, ncol = 5)

# now for each split, assess the onset and offset
participants_split1_train <- split_df %>%
    filter(split_number == 1 & group == "A") %>%
    pull(participant)
participants_split1_test <- split_df %>%
    filter(split_number == 1 & group == "B") %>%
    pull(participant)

# split1 data
summary_df_split1_train <- summary_df %>%
    filter(participant %in% participants_split1_train)

# fitting the GAM
meta_gam_split1_train <- brm(
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 10, by = condition) +
        (1 | participant),
    data = summary_df_split1_train,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4
    )
```

@@

@@

# computing and plotting bias, MAE, and variance
# bias := true_onset - median(sim_onset_distrib)
# MAE :=  mean(abs(sim_onset-true_onset))
# variance := variance(sim_onset_distrib)
sim_results %>%
    # keeping only the best brms results
    filter(eps == 0 & threshold == 20) %>%
    select(-eps,-threshold) %>%
    # filtering some alpha_level
    filter(alpha_level == 0.05) %>%
    select(-alpha_level) %>%
    pivot_longer(cols = -simulation_id, names_to = "measure", values_to = "value") %>%
    data.frame() %>%
    mutate(
        error = case_when(
            grepl("onset", measure) ~ abs(value - true_onset),
            grepl("offset", measure) ~ abs(value - true_offset)
            )
        ) %>%
    separate(measure, into = c("type", "method"), sep = "_", extra = "merge") %>%
    group_by(type, method) %>%
    summarise(
        MAE = median(error, na.rm = TRUE),
        variance = var(error, na.rm = TRUE),
        .groups = "drop"
        ) %>%
    ungroup() %>%
    mutate(type = factor(x = type, levels = c("onset", "offset") ) ) %>%
    mutate(
        method = factor(
            x = method,
            levels = c(
                "p", "bh", "by", "holm", "cluster_mass",
                "cluster_tfce", "cpt", "brms"
                ),
            labels = c(
                "Raw p-value", "FDR BH95", "FDR BY01", "Holm", "Cluster-mass",
                "TFCE", "Change point", "brms"
                ),
            )
        ) %>%
    ggplot(
        aes(
            x = MAE,
            y = variance,
            colour = method,
            fill = method
            )
        ) +
    geom_point(
        size = 2,
        pch = 21,
        colour = "white",
        show.legend = FALSE
        ) +
    geom_label_repel(
        aes(label = method),
        colour = "white",
        size = 3,
        segment.color = NA,
        show.legend = FALSE
        ) +
    scale_y_log10() +
    facet_wrap(~type) +
    scale_fill_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    scale_colour_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    labs(x = "Median absolute error (s)", y = "Variance (log-scale)")
```

```{r fig-simulation-distribution, echo = FALSE, fig.width = 9, fig.asp = 0.5, fig.cap = "Distributions of onset and offset estimates for each method."}
sim_results %>%
    # keeping only the best brms results
    filter(eps == 0 & threshold == 20) %>%
    # filtering some alpha_level
    filter(alpha_level == 0.05) %>%
    select(-alpha_level) %>%
    select(-eps,-threshold) %>%
    pivot_longer(
        cols = -simulation_id, 
        names_to = "measure",
        values_to = "value"
        ) %>%
    separate(measure, into = c("type", "method"), sep = "_", extra = "merge") %>%
    data.frame() %>%
    mutate(type = factor(x = type, levels = c("onset", "offset") ) ) %>%
    mutate(
        method = factor(
            x = method,
            levels = c(
                "p", "bh", "by", "holm", "cluster_mass",
                "cluster_tfce", "cpt", "brms"
                ),
            labels = c(
                "Raw p-value", "FDR BH95", "FDR BY01", "Holm",
                "Cluster-mass", "TFCE", "Change point", "brms"
                ),
            )
        ) %>%
    mutate(
        plot_row = ifelse(
            method %in% c("p-value", "FDR BH95", "FDR BY01", "Holm"), 0, 1
            )
        ) %>%
    ggplot(
        aes(
            x = value,
            colour = method,
            fill = method
            )
        ) +
    geom_vline(xintercept = true_onset) +
    geom_vline(xintercept = true_offset) +
    geom_density(aes(fill = NULL) ) +
    facet_grid(plot_row~type, scales = "free_x") +
    theme_light(base_size = 12, base_family = "Open Sans") +
    scale_fill_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    scale_colour_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    labs(x = "Onset/Offset (s)", y = "Density")
```

## Application to actual EEG data (reliability)

>>>>

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive processes. Typically, researchers are interested in testing whether and when such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics through time followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, while allowing efficient error-rates control, cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the precise onset or offset of effects (e.g., differences across conditions or groups). Here, we introduce a *model-based* approch for analysing M/EEG timeseriesn such as ERPs or decoding performance through time, and their differences across conditions or groups. This approach relies on Bayesian generalised additive multilevel models and outputs the posterior probabilility of the effect being above 0 (or above chance) at every timestep, while naturally taking into account the temporal dependencies and between-subject variability present in such data. Using both simulated and actual M/EEG data, we show that the proposed approach largely outperforms conventional methods in estimating the onset and offset of M/EEG effects, producing more precise and more reliable estimates. We provide an R package implementing the approach and illustrate how to integrate it into M/EEG statistical pipelines in MNE-Python."
keywords: [EEG, MEG, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

@@

## Problem statement

Description of cluster-based approaches [see @sassenhagen2019]... As pointed by the original authors themselves [@maris2007], "there There is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA \[false alarm\] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect" [@maris2007; @sassenhagen2019]... In contrast, the proposed approach, based on Bayesian generalised additive multilevel models, allows quantifying the probability of effects at the level of timesteps, voxels, sensors (etc), while naturally taking into account spatiotemporal dependencies.

@@

See also the rERP framework [@smith2014a; @smith2014b] and @tremblay2014...

From @dimigen2021: Recently, spline regression has been applied to ERPs [e.g., @hendrix2017; @kryuchkova2012]... GAMMs for EEG data [@abugaber2023; @meulman2015; @meulman2023]...

@@

Recently, @teichmann2022 provided a detailed tutorial on using Bayes factors (BFs) to analyse the 1D or 2D output from MVPA, that is, for testing, at every timestep, whether decoding performance is above chance level. However, this approach provides timeseries of BFs that ignores temporal dependencies..

@@

See for instance these tutorials [@sóskuthy2017; @winter2016] or application to phonetic data [@wieling2018; @sóskuthy2021] or this introduction [@baayen2020] or these reference books [@hastie2017; @wood2017]... application to pupillometry [@vanrij2019]... GAMLSS for neuroimaging data [@dinga2021]... Modelling auto-correlation in GAMMs + EEG example [@baayen2018]...

@@

$$
\begin{aligned} 
y_{i} &\sim \mathrm{EF}\left(\mu_{i}, \phi\right)\\
g\left(\mu_i\right) &= \underbrace{\mathbf{A}_{i} \gamma}_{\text{parametric part}} + \underbrace{\sum_{j=1}^{J} f_{j}\left(x_{ij}\right)}_{\text{non-parametric part}}
\end{aligned}
$$

where $y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)$ denotes that the observations $y_{i}$ are distributed as some member of the exponential family of distributions (e.g., Gaussian, Gamma, Beta, Poisson) with mean $\mu_{i}$ and scale parameter $\phi$; $g(\cdot)$ is the link function, $\mathbf{A}_{i}$ is the $i$th row of a known parametric model matrix, $\gamma$ is a vector of parameters for the parametric terms (to be estimated), $f_{j}$ is a smooth function of covariate $x_{j}$ (to be estimated as well). The smooth functions $f_{j}$ are represented in the model via penalised splines basis expansions of the covariates, that are a weighted sum of basis functions:

@@

Terminology: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., a cubic polynomial) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions.

@@

Now describe the Bayesian GAMM [@miller2025]... Proper inclusion of varying/random effects in the model specification protects against overly wiggly curves [@baayen2020]... Generalising to scale and shape or "distributional GAMs" [@rigby2005; @umlauf2018] and applied to neuroimaging data [@dinga2021]...

Instead of averaring, obtain the smooth ERP signal from multilevel GAM... less susceptible to outliers (Meulman et al., 2023)...

@@

Following the approach of @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see details and code in @yeung2004]. As in @rousselet_using_2025, the signal was generated from truncated Gaussian with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated such data for a group of 20 participants with 50 trials per participant and condition (@fig-eeg).

@@

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms` (i.e., weakly informative priors). We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects. @fig-post-prob-test shows the predictions of this model together with the raw data.

@@

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "cr", k = 20, by = condition),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/gam.rds"
    )
```

@@

# fitting the GAMM with factor smooth (per participant)
gamm <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition +
        s(time, bs = "cr", k = 20, by = condition) +
        s(time, participant, bs = "fs"),
    data = raw_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/gamm.rds"
    )
```

@@

Maybe better to use the p-MAP (i.e., density ratio between the null and the MAP)? And/or look at the derivative (e.g., <https://fromthebottomoftheheap.net/2014/05/15/identifying-periods-of-change-with-gams/>)?

```{r fig-post-prob-ratio, echo = FALSE, out.width = "75%", fig.asp = 0.75, fig.cap = "Ratio of posterior probability according to the GAM (on a log10 scale). Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.05
threshold <- 10

@@

# printing the results
# bind_rows(exceeding_times_gam, exceeding_times_gp) %>%
#     mutate(across(.cols = where(is.numeric), .fns = ~round(.x, 2) ) ) %>%
#     relocate(model, .before = cluster_onset)

```

@@

Next we fit a hierarchical/multilevel GAM using summary statistics of ERPs (mean and SD) at the participant level (similar to what is done in meta-analysis). Although it is possible to fit a GAMM at the single-trial level, we present a resource-efficient version of the model that is fitted directly on the by-participant summary statistics (mean and SD)...

@@

# fitting the GAM
meta_gam <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 20, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/meta_gam.rds"
    )
```

@@

Next, we assessed the performance of the proposed approach on actual MEG data (decoding results of Nalborczyk et al., in preparation). We conducted time-resolved multivariate pattern analysis (MVPA), also known as decoding... As a result, we have a timecourse of decoding performance (ROC AUC), bounded between 0 and 1, for each participant (for a total of 32 participants). Now, we want to *test* whether the group-level average decoding accuracy is above chance (i.e., 0.5) at each timestep (@fig-decoding-data). We fitted a similar GAM as discussed previously, but we replaced the $\mathrm{Normal}$ likelihood function by a $\mathrm{Beta}$ one to account for the bounded nature of AUC values (between 0 and 1) [for a tutorial on Beta regression, see @coretta2025]. Note that although we chose a basis dimension of $k=50$, which seems appropriate for the present data, this choice should be adapted according to the properties of the modelled data (e.g., signal-to-noise ratio, prior low-pass filtering, sampling rate, etc) and should be assessed by the usual model checking tools (e.g., posterior predictive checks). We also define a region of practical equivalence [ROPE, @kruschke2017], defined as the chance level plus the standard deviation of the (group-level average) decoding performance during the baseline period. This ensures that...[^1]

[^1]: Could an alternative approach be to include a predictor for "baseline vs. after baseline"?

```{r decoding-model, eval = FALSE}
# fitting the Beta GAM
meg_decoding_gam <- brm(
    auc ~ s(time, bs = "cr", k = 50),
    data = decoding_df,
    family = Beta(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4
    )
```

We assessed the reliability of the proposed approach using a form of permutation-based split-half reliability [e.g., @rosenblatt2018], which consisted of the following steps:

@@

```{r fig-decoding-data, echo = FALSE, out.width = "75%", fig.asp = 0.75, fig.cap = "Group-level average decoding performance (N=32) superimposed with the GAM predictions (in blue) and the region of practical equivalence (ROPE, in orange) computed from the baseline period. The blue horizontal line indicates the timesteps at which the posterior probability ratio is equal to or greater than 20."}
# importing the reshaped MEG decoding results
decoding_df <- read.csv(file = "code/decoding_results_reshaped.csv") %>%
    # removing some participants
    mutate(participant_id = cur_group_id(), .by = participant) %>%
    filter(participant_id < 33) %>%
    select(-participant_id)

# averaging across participants
decoding_summary_df <- decoding_df %>%
    summarise(
        auc_mean = mean(auc),
        auc_sd = sd(auc),
        .by = c(participant, time)
        )

# defining the chance level
chance_level <- 0.5

# computing the smallest effect size of interest (SESOI)
# as the SD of decoding performance during the baseline
sesoi <- decoding_summary_df %>%
    filter(time < 0) %>%
    summarise(auc_mean = mean(auc_mean), .by = time) %>%
    summarise(sesoi = sd(auc_mean) ) %>%
    # summarise(sesoi = sd(auc_mean) / sqrt(n_distinct(decoding_summary_df$participant) ) ) %>%
    pull(sesoi)

# fitting the Beta GAM
meg_decoding_gam <- brm(
    auc_mean ~ s(time, bs = "cr", k = 50),
    data = decoding_summary_df,
    family = Beta(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/meg_decoding_gam.rds"
    )

# some posterior predictive checks
# pp_check(gam_split, prefix = "ppd")
# pp_check(gam_split)
# pp_check(gam_split, type = "error_hist")
# pp_check(gam_split, type = "stat_2d")
# pp_check(gam_split, type = "ecdf_overlay")

# computing the posterior probability
prob_y_above_0 <- data.frame(time = unique(meg_decoding_gam$data$time) ) %>%
    # retrieving the posterior samples
    add_epred_draws(object = meg_decoding_gam) %>%
    # converting to dataframe
    data.frame() %>%
    # computing mean posterior probability at the group level
    group_by(time) %>%
    summarise(m = mean(.epred > (0 + chance_level + sesoi) ) ) %>%
    mutate(prob_ratio = m / (1 - m) ) %>%
    # ensuring there is no 0 or +Inf values
    mutate(prob_ratio = ifelse(is.infinite(prob_ratio), ndraws(meg_decoding_gam), prob_ratio) ) %>%
    mutate(prob_ratio = ifelse(prob_ratio == 0, 1 / ndraws(meg_decoding_gam), prob_ratio) ) %>%
    ungroup()

# finding onset and offset for the current threshold
# exceeding_times <- prob_y_above_0 %>%
#     dplyr::filter(prob_ratio > post_prob_ratio_threshold) %>%
#     summarise(
#         cluster_onset = min(time, na.rm = TRUE),
#         cluster_offset = max(time, na.rm = TRUE)
#         )
# or finding all clusters
# defining a function to find onset and offset in timeseries
# find_clusters <- function (mask, timeseries) {
#     
#     # initialising onset and offset values
#     onset <- NA
#     offset <- NA
#     
#     # identifying the onset and offset
#     masked_timesteps <- try(timeseries[which(mask)], silent = TRUE)
#     onset <- head(x = masked_timesteps, n = 1)
#     offset <- tail(x = masked_timesteps, n = 1)
#     
#     # returning the onset and offset
#     return (c(onset, offset) )
#     
# }
# 
# find_onset_offset(
#         tests_results$pval <= alpha_level,
#         tests_results$time
#         )
# prob_y_above_0 %>%
#     dplyr::filter(prob_ratio >= post_prob_ratio_threshold) %>%
#     dplyr::pull(time)
find_clusters <- function (df, threshold = 20) {

    df %>%
        mutate(above_threshold = prob_ratio >= threshold) %>%
        mutate(cluster_change = c(TRUE, diff(above_threshold) != 0) ) %>%
        filter(above_threshold) %>%
        mutate(cluster_id = cumsum(cluster_change)) %>%
        group_by(cluster_id) %>%
        summarise(cluster_onset = first(time), cluster_offset = last(time), .groups = "drop") %>%
        select(cluster_onset, cluster_offset)

post_prob_ratio_threshold <- 20
clusters <- find_clusters(prob_y_above_0, threshold = post_prob_ratio_threshold)

# checking model's predictions against raw data
plot(
    conditional_effects(x = meg_decoding_gam),
    line_args = list(colour = "steelblue", fill = "steelblue", alpha = 0.3),
    points = FALSE, plot = FALSE
    )[[1]] +
    geom_hline(yintercept = 0.5, linetype = 2) +
    geom_vline(xintercept = 0.0, linetype = 2) +
    annotate(
        geom = "rect",
        xmin = -Inf, xmax = Inf,
        ymin = chance_level - sesoi, ymax = chance_level + sesoi,
        fill ="orangered",
        alpha = 0.2
        ) +
    geom_segment(
        data = clusters,
        aes(
            x = cluster_onset,
            xend = cluster_offset,
            y = 0.48, yend = 0.48
            ),
        colour = "steelblue",
        inherit.aes = FALSE,
        lineend = "round", linewidth = 2
        ) +
    geom_line(
        data = decoding_summary_df %>% summarise(auc_mean = mean(auc_mean), .by = time),
        aes(x = time, y = auc_mean), inherit.aes = FALSE
        ) +
    labs(x = "Time (s)", y = "Decoding accuracy (ROC AUC)")
```

@@

This section is divided in two parts. First, we present the results from the simulation study, assessing the bias and variance of each method when applied to simulated data in which the ground truth is known. Second, we present the results obtained when applying the different methods to actual MEG data (decoding performance through time), assessing the reliability of each method and the stability of its estimates.

@@

# computing and plotting bias, MAE, and variance
# bias := true_onset - median(sim_onset_distrib)
# MAE :=  mean(abs(sim_onset-true_onset))
# variance := variance(sim_onset_distrib)
sim_results %>%
    # keeping only the best brms results
    filter(eps == 0 & threshold == 20) %>%
    select(-eps,-threshold) %>%
    # filtering some alpha_level
    filter(alpha_level == 0.05) %>%
    select(-alpha_level) %>%
    pivot_longer(cols = -simulation_id, names_to = "measure", values_to = "value") %>%
    data.frame() %>%
    mutate(
        error = case_when(
            grepl("onset", measure) ~ abs(value - true_onset),
            grepl("offset", measure) ~ abs(value - true_offset)
            # grepl("onset", measure) ~ value - true_onset,
            # grepl("offset", measure) ~ value - true_offset
            )
        ) %>%
    separate(measure, into = c("type", "method"), sep = "_", extra = "merge") %>%
    group_by(type, method) %>%
    summarise(
        MAE = median(error, na.rm = TRUE),
        variance = var(error, na.rm = TRUE),
        .groups = "drop"
        ) %>%
    ungroup() %>%
    mutate(type = factor(x = type, levels = c("onset", "offset") ) ) %>%
    mutate(
        method = factor(
            x = method,
            levels = c(
                "p", "bh", "by", "holm", "cluster_mass",
                "cluster_tfce", "cpt", "brms"
                ),
            labels = c(
                "Raw p-value", "FDR BH95", "FDR BY01", "Holm", "Cluster-mass",
                "TFCE", "Change point", "brms"
                ),
            )
        ) %>%
    ggplot(
        aes(
            x = MAE,
            y = variance,
            colour = method,
            fill = method
            )
        ) +
    # geom_vline(xintercept = 0, linetype = 2) +
    geom_point(
        size = 2,
        pch = 21,
        colour = "white",
        show.legend = FALSE
        ) +
    geom_label_repel(
        aes(label = method),
        colour = "white",
        size = 3,
        segment.color = NA,
        show.legend = FALSE
        ) +
    scale_y_log10() +
    facet_wrap(~type) +
    scale_fill_manual(values = met.brewer(name = "Johnson", n = 9) ) +
    scale_colour_manual(values = met.brewer(name = "Johnson", n = 9) ) +
    labs(x = "Median absolute error (s)", y = "Variance (log-scale)")
```
